A boundary-integral framework to simulate \\ viscous erosion of a porous medium

### 2018-04-09 docs/methods/erosion_methods.tex
parsed the tex code so there's no line breaks

\paragraph{Related Work}

In closely related work, Mitchell \& Spagnolie used a traction integral equation formulation to simulate viscous erosion of a single body in three dimensions \cite{mit-spa2016}. Their formulation was flexible enough to simulate a {\em mobile} body that erodes in various background flows (uniform flow, shear flow, etc.). An advantage of the traction formulation is that is directly provides the shear stress that is needed to evolve surfaces. 
In the present work, we aim to simulate the erosion of a porous medium and further we choose to fix the bodies to their initial locations. Thus, instead than simulating a single, mobile body as in \cite{mit-spa2016}, we will simulate the erosion of many, immobile bodies. Furthermore, we will employ a standard double-layer formulation, as it allows us to easily calculate bulk quantities such as vorticity, pressure, and velocity in a post-processing step.

@@

The layer potential~\eqref{eqn:genericLP} is approximated with the
trapezoid rule 
\begin{align}
  \phi(\xx) = \sum_{j=1}^N q_j K(\xx,\yy_j)\, ,
  \label{eqn:layerPotential}
\end{align}
where $q$ contains the quadrature weights, Jacobian, and density
function.  Based on numerical experiments, the error of the $N$-point
trapezoid rule is acceptable for points $\xx$ that are more than five 
arclength spacings from $\bd\Omega$. For points that are closer than
five arclength spacings to $\bd\Omega$, we use an upsampled trapezoid
rule.  The required upsampling of the geometry and the density function
is done with a Fourier interpolant.  To minimize the number of
discretization points that are used to approximate the layer
potential~\eqref{eqn:layerPotential}, we use the same upsampling rate
for all target points in dyadic intervals (Table~\ref{tbl:upsampling}).
For points that are closer than $5/16$ of an arclength from $\bd\Omega$,
a value of 0 is simply assigned to $\phi$.
\newcolumntype{V}{>{\centering\arraybackslash}m{4em}}
\newcolumntype{N}{@{}m{0pt}@{}}
\begin{table}[htpb]
\centering
%\begin{tabular}{|c|ccccc|}
\begin{tabular}{|V|V|V|V|V|V|N}
  \hline
  $d(\xx,\bd\Omega)$ &
  $\left[5h,\infty\right]$ &
  $\left[\frac{5}{2}h,5h\right]$ &
  $\left[\frac{5}{4}h,\frac{5}{2}h\right]$ & 
  $\left[\frac{5}{8}h,\frac{5}{4}h\right]$ &
  $\left[\frac{5}{16}h,\frac{5}{8}h\right]$ & \\ [2ex] 
  \hline
  Upsample Rate & 1 & 2 & 4 & 8 & 16 & \\
  \hline
\end{tabular}
\caption{\label{tbl:upsampling}The upsampling rate of our near-singular
integration scheme.  $d(\xx,\bd\Omega)$ is the distance between a point
$\xx \in \Omega$ and the boundary of the domain $\bd\Omega$, and $h$ is
an arclength spacing.}
\end{table}

@@

	We show in Fig.~\ref{fixed_intface}(a) several interfaces of this modified erosion process, with the same spatial resolution, $N_\iin = 1024$, as before. Notice the body obtains a similar profile as observed previously, but now with fixed area. Importantly, more numerical accuracy is available to quantify the terminal state. The inset in Fig.~\ref{fixed_intface}(a) zooms on the rear corner demonstrate that, due to our numerical regularization, the body is actually smooth on a fine scale. We show in Fig.~\ref{fixed_intface}(b) the stress distributions on the same interfaces. As observed previously, the stress becomes more evenly distributed as erosion reshapes the body. The slight rise in shear near the front/rear corners that was observed in Fig.~\ref{shrink_intface}(b) is no longer present, suggesting that this effect was due to loss of accuracy. Rather, the late-time stress distributions in Fig.~\ref{fixed_intface}(b) decrease monotonically to zero near the stagnation points. In this figure, time is normalized by the vanishing time $t_f = 1.79 \times 10^{-2}$ from the shrinking simulations (i.e.~Fig.~\ref{shrink_intface}). Curiously, about 3 multiples of $t_f$ are required to reach a visible steady state, which contrasts with previous studies on high-Reynolds-number erosion in which a steady state usually emerged before the body disappeared~\cite{moo-ris-chi-zha-she2013}.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.85 \textwidth]{./figs/arangle.pdf}
\caption{Shape characteristics of an eroding body: (a) The aspect ratio (width to height) versus time. (b) The measured opening angle versus time. For both measurements, we show three cases of the smoothing parameters: $\eps = \sigma = 20/1024, 10/1024,$ and 5/1024. In the finest simulation ($5/1024$) the aspect ratio converges to 2.7 and the opening angle converges to the predicted value of $102^{\circ}$ in long time.}
\label{fig:arangle}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure from datafile 01circ1024c5
% Uses 01circ1024.in with epsfac = 5, sigfac = 5, dt = 2e-4, fixarea = 1, fixpdrop = 0

@@

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}%[htbp]
\begin{center}
\caption{The long-time aspect ratio and opening-angle measurements for three...
} 
\vspace{0.3 pc}
\label{table:arangle}
\begin{tabular}{c c c}
\hline
\hspace{0.5pc} smoothing parameters: $\eps$ and $\sigma$
\hspace{0.5pc} & final aspect ratio 
\hspace{0.5pc} & final opening angle \\
\hline
20/1024		& 2.55	& $110^\circ \pm 6^\circ$	\\
10/1024		& 2.62	& $104^\circ \pm 4^\circ$	\\
5/1024		& 2.65	& $102^\circ \pm 2^\circ$	\\
\hline
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% npts = 1024

@@

\subsection{Convergence test}

As a final validation, we now verify that our method converges with the
expected second-order accuracy in time. For this, we run the same
single-body simulation shown in Figs.~\ref{01bodseq} and~\ref{shrink_intface} (except with a coarser $N_\iin=256$) to a stopping time of $t_s = 10^{-2}$. This stopping time is 56\% of the vanishing time, and was selected so that the body will have eroded substantially but is not small enough to compromise accuracy (due to the stress diverging).
We assess error by measuring the $L^2$-difference in the body shape obtained from simulations of successive resolution, with $\Delta t$ halved at each stage. In Table~\ref{convtab}, we report these errors along with the order of accuracy, obtained by Richardson-extrapolation-type calculations. The table confirms that our method converges with expected second-order accuracy. The last column of the table shows the CPU time taken for the simulation to run on a single-processor machine, demonstrating a nearly linear increase in computational time. Though the main parts of the algorithm are easily paralyzed, the single-processor times provide a simple assessment of method's complexity.

@@

% Table: GMRES iterations
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}%[htbp]
\begin{center}
\caption{The number of GMRES iterations required to reach a fixed
tolerance of $10^{-}8$ in our simulations. The left side shows a case of fixed geometry, 10 circular bodies, with increasing spatial resolution, $N_\iin$, demonstrating that the iteration count is independent of resolution. In the right side, we fix the resolution and sequentially increase the number of bodies (all circular) from 10 to 50. The iteration count increases roughly linearly with the number of bodies.
}
\vspace{0.3 pc}
\label{itertab}
\begin{tabular}{c c | c c}
\hline
\hspace{0.5pc} Resolution, $N_{\iin}$
\hspace{0.5pc} & GMRES iterations 
\hspace{0.5pc} &  Number of bodies
\hspace{0.5pc} & GMRES iterations  \\
\hspace{0.0pc} (10 bodies) &
\hspace{0.5pc} & ($N_\iin = 128$ fixed) & \\
\hline
%
32	& 228	& 10 	& 223	\\
64    	& 226	& 15 	& 292	\\
128	& 223	& 20 	& 363	\\
256	& 223	& 30 	& 496	\\
512	& 223	& 40 	& 676	\\
1024	& 223	& 50 	& 991	\\
%
\hline
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

>>>>

\paragraph{Related Work}
In closely related work, Mitchell \& Spagnolie used a traction integral equation formulation to simulate viscous erosion of a single body in three dimensions \cite{mit-spa2016}. Their formulation was flexible enough to simulate a {\em mobile} body that erodes in various background flows (uniform flow, shear flow, etc.). An advantage of the traction formulation is that is directly provides the shear stress that is needed to evolve surfaces.  In the present work, we aim to simulate the erosion of a porous medium and further we choose to fix the bodies to their initial locations. Thus, instead than simulating a single, mobile body as in \cite{mit-spa2016}, we will simulate the erosion of many, immobile bodies. Furthermore, we will employ a standard double-layer formulation, as it allows us to easily calculate bulk quantities such as vorticity, pressure, and velocity in a post-processing step.

@@

The layer potential~\eqref{eqn:genericLP} is approximated with the trapezoid rule 
\begin{align}
  \phi(\xx) = \sum_{j=1}^N q_j K(\xx,\yy_j)\, ,
  \label{eqn:layerPotential}
\end{align}
where $q$ contains the quadrature weights, Jacobian, and density function.  Based on numerical experiments, the error of the $N$-point trapezoid rule is acceptable for points $\xx$ that are more than five arclength spacings from $\bd\Omega$. For points that are closer than five arclength spacings to $\bd\Omega$, we use an upsampled trapezoid rule.  The required upsampling of the geometry and the density function is done with a Fourier interpolant.  To minimize the number of discretization points that are used to approximate the layer potential~\eqref{eqn:layerPotential}, we use the same upsampling rate for all target points in dyadic intervals (Table~\ref{tbl:upsampling}).  For points that are closer than $5/16$ of an arclength from $\bd\Omega$, a value of 0 is simply assigned to $\phi$.
\newcolumntype{V}{>{\centering\arraybackslash}m{4em}}
\newcolumntype{N}{@{}m{0pt}@{}}
\begin{table}[htpb]
\centering
\begin{tabular}{|V|V|V|V|V|V|N}
  \hline
  $d(\xx,\bd\Omega)$ &
  $\left[5h,\infty\right]$ &
  $\left[\frac{5}{2}h,5h\right]$ &
  $\left[\frac{5}{4}h,\frac{5}{2}h\right]$ & 
  $\left[\frac{5}{8}h,\frac{5}{4}h\right]$ &
  $\left[\frac{5}{16}h,\frac{5}{8}h\right]$ & \\ [2ex] 
  \hline
  Upsample Rate & 1 & 2 & 4 & 8 & 16 & \\
  \hline
\end{tabular}
\caption{\label{tbl:upsampling}The upsampling rate of our near-singular integration scheme.  $d(\xx,\bd\Omega)$ is the distance between a point $\xx \in \Omega$ and the boundary of the domain $\bd\Omega$, and $h$ is an arclength spacing.}
\end{table}

@@

We show in Fig.~\ref{fixed_intface}(a) several interfaces of this modified erosion process, with the same spatial resolution, $N_\iin = 1024$, as before. Notice the body obtains a similar profile as observed previously, but now with fixed area. Importantly, more numerical accuracy is available to quantify the terminal state. The inset in Fig.~\ref{fixed_intface}(a) zooms on the rear corner demonstrate that, due to our numerical regularization, the body is actually smooth on a fine scale. We show in Fig.~\ref{fixed_intface}(b) the stress distributions on the same interfaces. As observed previously, the stress becomes more evenly distributed as erosion reshapes the body. The slight rise in shear near the front/rear corners that was observed in Fig.~\ref{shrink_intface}(b) is no longer present, suggesting that this effect was due to loss of accuracy. Rather, the late-time stress distributions in Fig.~\ref{fixed_intface}(b) decrease monotonically to zero near the stagnation points. In this figure, time is normalized by the vanishing time $t_f = 1.79 \times 10^{-2}$ from the shrinking simulations (i.e.~Fig.~\ref{shrink_intface}). Curiously, about 3 multiples of $t_f$ are required to reach a visible steady state, which contrasts with previous studies on high-Reynolds-number erosion in which a steady state usually emerged before the body disappeared~\cite{moo-ris-chi-zha-she2013}.  
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.85 \textwidth]{./figs/arangle.pdf}
\caption{Shape characteristics of an eroding body: (a) The aspect ratio (width to height) versus time. (b) The measured opening angle versus time. For both measurements, we show three cases of the smoothing parameters: $\eps = \sigma = 20/1024, 10/1024,$ and 5/1024. In the finest simulation ($5/1024$) the aspect ratio converges to 2.7 and the opening angle converges to the predicted value of $102^{\circ}$ in long time.}
\label{fig:arangle}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure from datafile 01circ1024c5
% Uses 01circ1024.in with epsfac = 5, sigfac = 5, dt = 2e-4, fixarea = 1, fixpdrop = 0

@@

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}%[htbp]
\begin{center}
\todo[inline]{finish caption (just didn't want to forget it)}
\caption{The long-time aspect ratio and opening-angle measurements for three...
} 
\vspace{0.3 pc}
\label{table:arangle}
\begin{tabular}{c c c}
\hline
\hspace{0.5pc} smoothing parameters: $\eps$ and $\sigma$
\hspace{0.5pc} & final aspect ratio 
\hspace{0.5pc} & final opening angle \\
\hline
20/1024		& 2.55	& $110^\circ \pm 6^\circ$	\\
10/1024		& 2.62	& $104^\circ \pm 4^\circ$	\\
5/1024		& 2.65	& $102^\circ \pm 2^\circ$	\\
\hline
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% npts = 1024

@@

\subsection{Convergence test}
As a final validation, we now verify that our method converges with the expected second-order accuracy in time. For this, we run the same single-body simulation shown in Figs.~\ref{01bodseq} and~\ref{shrink_intface} (except with a coarser $N_\iin=256$) to a stopping time of $t_s = 10^{-2}$. This stopping time is 56\% of the vanishing time, and was selected so that the body will have eroded substantially but is not small enough to compromise accuracy (due to the stress diverging).  We assess error by measuring the $L^2$-difference in the body shape obtained from simulations of successive resolution, with $\Delta t$ halved at each stage. In Table~\ref{convtab}, we report these errors along with the order of accuracy, obtained by Richardson-extrapolation-type calculations. The table confirms that our method converges with expected second-order accuracy. The last column of the table shows the CPU time taken for the simulation to run on a single-processor machine, demonstrating a nearly linear increase in computational time. Though the main parts of the algorithm are easily paralyzed, the single-processor times provide a simple assessment of method's complexity.

@@

% Table: GMRES iterations
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}%[htbp]
\begin{center}
\caption{The number of GMRES iterations required to reach a fixed tolerance of $10^{-8}$ in our simulations. The left side shows a case of fixed geometry, 10 circular bodies, with increasing spatial resolution, $N_\iin$, demonstrating that the iteration count is independent of resolution. In the right side, we fix the resolution and sequentially increase the number of bodies (all circular) from 10 to 50. The iteration count increases roughly linearly with the number of bodies.  }
\vspace{0.3 pc}
\label{itertab}
\begin{tabular}{c c | c c}
\hline
\hspace{0.5pc} Resolution, $N_{\iin}$
\hspace{0.5pc} & GMRES iterations 
\hspace{0.5pc} &  Number of bodies
\hspace{0.5pc} & GMRES iterations  \\
\hspace{0.0pc} (10 bodies) &
\hspace{0.5pc} & ($N_\iin = 128$ fixed) & \\
\hline
%
32	& 228	& 10 	& 223	\\
64    	& 226	& 15 	& 292	\\
128	& 223	& 20 	& 363	\\
256	& 223	& 30 	& 496	\\
512	& 223	& 40 	& 676	\\
1024	& 223	& 50 	& 991	\\
%
\hline
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
